Determinism, Complexity, and Predictability in Computer Performance

### 2014-08-04 response.tex

This short paragraph in this paper, which explains the efficiencies
and inefficiencies of these schemes, was intended to serve two
purposes: first, to explain why we have included these simple
prediction schemes and not just the more-complicated ones; second, to
help in explaining the specific results described later in the paper.

\smallskip

\emph{9) A better justification of the prediction error metrics could
  be also in order if this work is continued. At least, comparison of
  prediction errors quantified with different metrics seems
  necessary. Especially, taking into account the following major
  concerns.}

We have extended the justification for our use of the MASE metric; see
the last paragraph in Section VI.  As mentioned there, the paper in
which MASE was introduced---\alert{[[our reference [21]]]}---includes
an extensive evaluation of this metric, as well as a comprehensive
comparison to other metrics.  If the reviewer thinks that it would
strengthen our paper, we can recapitulate more of that material in our
paper.  As part of our current work on this topic, we are evaluating
how other choices of metric affect the shape of our curves, but we
feel that a comprehensive study of this issue is best left as a sequel
to this work.

@@

The WPE vs. MASE points of the transformed data move---upwards,
generally---but they remain within the shaded region.  Indeed, the
values for the transformed version of {\tt dgesdd}$_5$ (dark-green
icons) move {\sl closer} to the curve.

This series of experiments supports our claims of generality: a
transformation like this effectively changes the generating process,
but does not affect the performance of our model-free heuristic.  And
the curves that capture that heuristic---which were derived from data
from a single computer-performance experiment---do not only hold for
traces from that experiment and transformed versions of them.  The
preliminary results of our broader study, currently underway, indicate
that they also hold for data from synthetic examples (the H\'{e}non
map, the Lorenz equations, and the output of a purely stochastic
random-walk process) and for real-world data from a different system
(SFI data set A).  The image below shows these results:

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras_with_nonlinear_all_points}
\end{center}

The blue fit lines here are derived from the original 120 traces; the
magenta one is fit to those traces plus the nonlinear transformations
of {\tt dgesdd}$_2$, {\tt dgesdd}$_5$, and {\tt dgesdd}$_6$, the $x$
coordinate of a 10,000-point orbit from the H\'{e}non map with $a=1.4$
and $b=0.3$, the $x$ coordinate of a 10,000-step 4th-order Runge-Kutta
solution of the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$
and $\Delta t=0.01$, a 10,000-step random-walk time series generated
by taking the cumulative sum of $\mu=0$, $\sigma=1$ \alert{[[generated
      how?  Gaussian random variables,]]} and data set ``A'' from the
Santa Fe Institute time-series competition website.

### 2014-08-04 response.tex

Note that this fit ($y = a \log(b x + 1)$) is different than in the
submitted version.  A careful examination of the data and some careful
thought about the meaning and theoretical bounds of MASE---prompted by
the reviewer's questions---suggested that this was a better match to
the data.  This analysis and consideration is chronicled in the third
paragraph of the results section.  Finally, note that there are a lot
more points on this plot than in our original version; this is
discussed further below.

\smallskip

\emph{1) Introduction is too long, but the main problem is not
  formulated definitely enough. I see that the main point is the
  approximately linear (or at least one-to-one) relationship between
  WPE and the best prediction error. This aim is not claimed
  explicitly in the Introduction. All other aspects of the manuscript
  do not seem important enough to be published in a separate article.}

We have shortened the introduction and modified the manuscript
throughout to make it more clear that our goal was to derive a
practical criterion for best-case prediction error in a model-free
way, and to test it in the context of empirical data.  We have also
changed the title and abstract to line up with this goal.

@@

Regarding the first part of this point: while PE is invariant under an
invertible nonlinear transformation of the values in the time series,
WPE is not---which amplifies the potential concerns about this issue.
To explore this, we performed the invertible nonlinear transformation
$f(x)=5e^{x(t)}$ on several of our traces ($x(t)=$ {\tt dgesdd}$_2$,
{\tt dgesdd}$_5$, and {\tt dgesdd}$_6$).  The image below shows WPE
vs. MASE results of the best predictor on these three traces, along
with those on the untransformed originals:

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/nonlinear_transform_data}
\end{center}

The WPE vs. MASE points of the transformed data move---upwards,
generally---but they remain within the shaded region.  Indeed, the
values for the transformed version of {\tt dgesdd}$_5$ (dark-green
icons) move {\sl closer} to the curve.

This series of experiments supports our claims of generality: a
transformation like this effectively changes the generating process,
but does not affect the performance of our model-free heuristic.  And
the curves that capture that heuristic---which were derived from data
from a single computer-performance experiment---do not only hold for
traces from that experiment and transformed versions of them.  The
preliminary results of our broader study, currently underway, indicate
that they also hold for data from synthetic examples and for
real-world data from a different system.  The image below shows some
of these results:

@@

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras_with_nonlinear_all_points}
\end{center}

The blue fit lines here are derived from the original 120 traces; the
magenta one is fit to those traces plus the nonlinear transformations
of {\tt dgesdd}$_2$, {\tt dgesdd}$_5$, and {\tt dgesdd}$_6$, the $x$
coordinate of a 10,000-point orbit from the H\'{e}non map with $a=1.4$
and $b=0.3$, the $x$ coordinate of a 10,000-step 4th-order Runge-Kutta
solution of the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$
and $\Delta t=0.01$, a 10,000-step random-walk time series generated
by taking the cumulative sum of $\mu=0$, $\sigma=1$ \alert{[[generated
      how?  Gaussian random variables,]]} and data set ``A'' from the
Santa Fe Institute time-series competition website.

Note that the inclusion of these additional 49 points hardly changed
the main central curve of the fit.  Indeed, doing so caused the
one-$\sigma$ volume of the fit to shrink (viz., the dashed magenta
curves inside the dashed blue curves).  \emph{That is, the additional
  points firmed up the findings.}  The magenta curve reflects a map, a
flow, a purely stochastic signal, a variety of real-world data from
two different systems (laser, computer), and nonlinear transformations
of the computer-performance data---again, supporting the generality of our
heuristic.

### 2014-08-04 concl.tex
adding some comments

These curves were determined empirically using a specific error metric
and a finite set of forecast methods and time series traces.  If one
uses a different error metric, the geometry---and the heuristic---will
no longer apply.  And while the methods and traces were chosen to be
representative of the practice, they are of course not completely
comprehensive.  It is certainly possible, for instance, that the
nonlinear dynamics of computer performance is subtly different from
the nonlinear dynamics of other systems.  Our preliminary results on
other systems (Henon, Lorenz, white noise, SFI ``A'', transformations
of the computer performance data) lead us to believe that our results
will generalize beyond the examples described in this paper.  We are
in the process of following up on that exploration with a broader
study of data, forecast methods, and error metrics.

\alert{I'm not sure this paragraph adds anything.  Can/should we
  remove it?  Given that information is a fundamental limitation in
  predictability, then gathering and using more information is an
  obvious next step.  But there is an equally obvious tension here
  between data length and prediction speed: a forecast that requires
  half a second to compute is not useful for the purposes of real-time
  control of a computer system with a MHz clock rate, for instance.
  Another alternative is to sample several system variables
  simultaneously and build multivariate models.  This is a particular
  challenge in nonlinear LMA-type models, since multivariate
  delay-coordinate embedding (e.g.,
  \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
  computationally prohibitive.  We are working on alternative methods
  that sidestep that complexity.}

>>>>

These curves were determined empirically using a specific error metric
and a finite set of forecast methods and time series traces.  If one
uses a different error metric, the geometry---and the heuristic---will
no longer apply.  And while the methods and traces were chosen to be
representative of the practice, they are of course not completely
comprehensive.  It is certainly possible, for instance, that the
nonlinear dynamics of computer performance is subtly different from
the nonlinear dynamics of other systems.  Our preliminary results on
other systems (H\'enon, Lorenz, white noise, SFI ``A'', transformations
of the computer performance data) lead us to believe that our results
will generalize beyond the examples described in this paper.  We are
in the process of following up on that exploration with a broader
study of data, forecast methods, and error metrics.

{\color{blue}agreed, I don't think it addds anything}\alert{I'm not sure this paragraph adds anything.  Can/should we
  remove it?  Given that information is a fundamental limitation in
  predictability, then gathering and using more information is an
  obvious next step.  But there is an equally obvious tension here
  between data length and prediction speed: a forecast that requires
  half a second to compute is not useful for the purposes of real-time
  control of a computer system with a MHz clock rate, for instance.
  Another alternative is to sample several system variables
  simultaneously and build multivariate models.  This is a particular
  challenge in nonlinear LMA-type models, since multivariate
  delay-coordinate embedding (e.g.,
  \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
  computationally prohibitive.  We are working on alternative methods
  that sidestep that complexity.}

### 2014-08-04 related.tex
adding some comments

Modeling time-series data for the purposes of prediction dates back at
least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
then, hundreds, if not thousands, of strategies have been developed
for a wide variety of prediction tasks.  The purpose of this paper is
not to add a new weapon to this arsenal, nor to do any sort of
theoretical assessment or comparison of existing methods.  Our goals
are focused more on the \emph{practice} of prediction: \emph{(i)} to
empirically quantify the predictive structure that is present in a
real-valued scalar time series and \emph{(ii)} to explore how the
performance of prediction methods is related to that inherent
complexity.  It would, of course, be neither practical nor interesting
to report results for every existing forecast method; instead, we
choose a representative set, as described in Section~\ref{sec:model}.

>>>>

Modeling time-series data for the purposes of prediction dates back at
least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
then, hundreds, if not thousands, of strategies have been developed
for a wide variety of prediction tasks.  The purpose of this paper is
not to add a new weapon to this arsenal, nor to do any sort of
theoretical assessment or comparison of existing methods.  Our goals
are focused more on the \emph{practice} of prediction: \emph{(i)} \alert{to
empirically quantify the predictive structure that is present in a
real-valued scalar time series} and \emph{(ii)} to explore how the
performance of prediction methods is related to that inherent
complexity.  It would, of course, be neither practical nor interesting
to report results for every existing forecast method; instead, we
choose a representative set, as described in Section~\ref{sec:model}.

### 2014-08-04 results.tex
adding some comments

The curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from finite sets of
methods and data.  We put a lot of thought and effort into making
these sets representative and comprehensive.  The forecast methods
involved range from the simple to the sophisticated; the time-series
data analyzed in this section was sampled from a system whose behavior
spans the dynamical behavior space.  While we are cautiously
optimistic about the generality of our conclusions, more exploration
will be required before we can make any such claim.  Our preliminary
work along those lines shows that data from the Henon map
\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set
\cite{weigend-book,sfi-data}, and a white-noise process all fall
within the one-$\sigma$ volume of the fit in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
region, as do various nonlinear transformations of \svdtwo, \svdfive
and \svdsix.

>>>>

The curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from finite sets of
methods and data.  We put a lot of thought and effort into making
these sets representative and comprehensive.  The forecast methods
involved range from the simple to the sophisticated; the time-series
data analyzed in this section was sampled from a system whose behavior
spans the dynamical behavior space.  While we are cautiously
optimistic about the generality of our conclusions, more exploration
will be required before we can make any such claim.  Our preliminary
work along those lines shows that data from the H\'enon map
\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set\footnote{\alert{This dataset can be retrieved from \url{http://www-psych.stanford.edu/~andreas/Time-Series/SantaFe.html}.}}
\cite{weigend-book,sfi-data},\alert{maybe make the website a footnote instead} and a random-walk process all fall
within the one-$\sigma$ volume of the fit in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
region, as do various nonlinear transformations of \svdtwo, \svdfive
and \svdsix.
